---
title: "Data Visualization in R"
subtitle: "Week 8, Part 1"
author: "Jeff Jacobs"
institute: "DSAN Summer Bootcamp 2024<br>Georgetown University"
format:
  revealjs:
    echo: true
    code-fold: false
    slide-number: true
    df-print: kable
    html-math-method: mathjax
    scrollable: true
    footer: "2024 DSAN Bootcamp W08-1: Visualization"
    theme: ["../_jjslides.scss"]
    output-file: "slides.html"
    include-in-header:
      text: "<link href='https://cdn.jsdelivr.net/npm/bootstrap@5.3.3/dist/css/bootstrap.min.css' rel='stylesheet' integrity='sha384-QWTKZyjpPEjISv5WaRU9OFeRpok6YctnYmDr5pNlyT2bRjXh0JMhjY6hW+ALEwIH' crossorigin='anonymous'><link rel='stylesheet' href='https://cdn.jsdelivr.net/npm/bootstrap-icons@1.11.3/font/bootstrap-icons.min.css'><script src='https://cdn.jsdelivr.net/npm/bootstrap@5.3.3/dist/js/bootstrap.bundle.min.js' integrity='sha384-YvpcrYf0tY3lHB60NNkmXc5s9fDVZLESaAA55NDzOxhy9GkcIdslK1eN7N6jIeHz' crossorigin='anonymous'></script>"
    simplemenu:
      flat: true
      barhtml:
        header: "<div class='menubar'><span style='position: absolute; left: 8; padding-left: 8px;'><a href='./index.html'>&larr; Return to Notes</a></span><ul class='menu'></ul></div>"
      scale: 0.5
    revealjs-plugins:
      - simplemenu
  html:
    df-print: kable
    html-math-format: mathjax
    echo: true
    code-fold: true
    output-file: "index.html"
    include-in-header:
      text: "<link href='https://cdn.jsdelivr.net/npm/bootstrap@5.3.3/dist/css/bootstrap.min.css' rel='stylesheet' integrity='sha384-QWTKZyjpPEjISv5WaRU9OFeRpok6YctnYmDr5pNlyT2bRjXh0JMhjY6hW+ALEwIH' crossorigin='anonymous'><link rel='stylesheet' href='https://cdn.jsdelivr.net/npm/bootstrap-icons@1.11.3/font/bootstrap-icons.min.css'><script src='https://cdn.jsdelivr.net/npm/bootstrap@5.3.3/dist/js/bootstrap.bundle.min.js' integrity='sha384-YvpcrYf0tY3lHB60NNkmXc5s9fDVZLESaAA55NDzOxhy9GkcIdslK1eN7N6jIeHz' crossorigin='anonymous'></script>"
---

::: {.content-visible unless-format="revealjs"}

<center>
<a class="h2" href="./slides.html" target="_blank">Open slides in new window &rarr;</a>
</center>

:::

# Why Are We *Starting* with Tidyverse Here? {data-stack-name="Overview" .crunch-ul}

## Base R vs. Tidyverse {.smaller .crunch-details}

::: {.columns}
::: {.column width="50%"}

```{r}
#| label: base-r-plot
#| code-fold: true
#| fig-height: 8
source("../_globals.r")
lexp_df <- read.csv("assets/lexp_by_ward.csv")
plot(
  lexp_df$ward, lexp_df$life_exp,
  main="DC Life Expectancy by Ward (2015)",
  xlab="Ward", ylab="Life Expectancy",
  pch=19, cex.lab=2, cex.main=2, cex.axis=2
)
```

:::
::: {.column width="50%"}

```{r}
#| label: ggplot-ward-lollipop
#| code-fold: true
#| fig-height: 8
#| fig-cap: "Data from [DC Dept. of Health](https://disb.dc.gov/sites/default/files/dc/sites/disb/publication/attachments/her_summary_report_final_with_letter_and_table_02_08_2019.pdf)"
library(tidyverse)
us_mean <- 78.69
lexp_df <- read_csv("assets/lexp_by_ward.csv")
lexp_df <- lexp_df |> mutate(
  above_avg = ifelse(life_exp > us_mean, "Above Mean", "Below Mean")
)
lexp_df |> ggplot(aes(x=as.factor(ward), y=life_exp, color=above_avg)) +
  geom_point(size=5) +
  geom_segment(
    aes(xend=ward, yend=us_mean),
    linewidth=1.5
  ) +
  geom_hline(
    aes(
      yintercept=us_mean,
      linetype="National Mean"
    ),
    linewidth=1
  ) +
  dsan_theme("half") +
  theme(
    plot.title = element_text(hjust = 0.5)
  ) +
  labs(
    x = "Ward",
    y = "Life Expectancy",
    title = "DC Life Expectancy by Ward (2015)"
  ) +
  scale_linetype_manual("test", values=c("dashed")) +
  remove_legend_title()
```

:::
:::

## Plus Way Wider Range of Plot Types!

(All using the same syntax)


```{r}
#| label: ward-choropleth
#| code-fold: true
#| fig-cap: "Shapefile from [OpenData.DC.gov](https://opendata.dc.gov/datasets/c5cd8b40fb784548a6680aead5f919ed/explore)"
library(sf)
my_sf <- read_sf("assets/Wards_from_2022.geojson")
my_sf_merged <- my_sf |> left_join(
  lexp_df, by=c("WARD"="ward")
)
my_sf_merged |> ggplot(aes(fill=life_exp)) +
  geom_sf() +
  dsan_theme() +
  theme(
    axis.line = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  ) +
  labs(
    title = "DC Life Expectancy by Ward (2015)",
    fill = "Life Exp."
  ) +
  scale_fill_viridis_c()
```

# Grammar of Graphics {data-stack-name="Grammar of Graphics"}

## Intuition

* **Sentences** are made up of *subjects*, *verbs*, *objects* (grammatical structure)
* **Visualizations** are made up of *encodings*, *coordinate systems*, *aesthetic mappings*

